---
title: "Annual institutional sports equity report"
format: pdf
theme: litera
toc: FALSE
fontsize: 12pt
embed-resources: TRUE
page-layout: full
params:
  inst: Furman University 
  year: 2021
geometry:
      - top=2cm
---

```{r}
#| echo: false
#| output: false

library(tidyverse)
library(lubridate)
library(assertr)
library(quarto)
library(readxl)
library(glue)
library(kableExtra)
library(janitor)

curr_year <- params$year
inst      <- params$inst

root <- getwd()

dta <- read_excel(file.path(root,glue("EADA_{curr_year}.xlsx")), sheet = 1) %>%
  filter(institution_name == inst) %>%
  verify(nrow(.) == 1)

location <- paste0(dta %>% pull(city_txt),", ", dta %>% pull(state_cd))

sy <- paste0(params$year - 1,"-",params$year)

```

```{=latex}
\vspace{-2.5cm}
```

**Year:** `r sy`

**Institution:** `r dta %>% pull(institution_name)`

**Location:** `r location`

**Institution type:** `r dta %>% pull(sector_name)`

**NCAA classification:** `r dta %>% pull(classification_name)`

**Total student body size:** `r format(dta %>% pull(EFTotalCount), big.mark = ",")`

## I. Summary of sports offerings and participation

```{r}
#| echo: false
#| output: true

all_sports <- dta %>%
  select(starts_with(c("PARTIC_WOMEN","PARTIC_MEN"))) %>%
  pivot_longer(cols = everything(), 
               names_to = c("gender", "Sport"), 
               values_to = "participants",
               names_pattern = "PARTIC_(\\w+)_(.*)") %>%
  filter(!is.na(participants) & participants > 0) %>%
  mutate(gender = str_to_title(gender),
         Sport  = case_when(Sport == "BchVoll"    ~ "Beach Volleyball",
                            Sport == "Bskball"    ~ "Basketball",
                            Sport == "Eqstrian"   ~ "Equestrian",
                            Sport == "FldHcky"    ~ "Field Hockey",
                            Sport == "Gymn"       ~ "Gymnastics",
                            Sport == "IceHcky"    ~ "Ice Hockey",
                            Sport == "Lacrsse"    ~ "Lacrosse",
                            Sport == "OthSpts"    ~ "other Sports",
                            Sport == "SwimDivng"  ~ "Swimming \\& Diving",
                            Sport == "SynSwim"    ~ "Synchronized Swimming",
                            Sport == "TblTennis"  ~ "Table Tennis",
                            Sport == "Trckcomb"   ~ "Track \\& Field (Combined)",
                            Sport == "TrkFldIn"   ~ "Indoor Track \\& Field",
                            Sport == "TrkFldOut"  ~ "Outdoor Track \\& Field",
                            Sport == "Vollball"   ~ "Volleyball",
                            Sport == "WaterPolo"  ~ "Water Polo",
                            Sport == "WgtLift"    ~ "Weightlifting",
                            Sport == "XCountry"   ~ "Cross country",
                            TRUE                  ~ Sport))

total_sports <- all_sports %>%
  group_by(gender) %>%
  summarise(n_sports = n()) %>%
  pivot_wider(names_from = gender, values_from = n_sports) %>%
  ungroup() %>%
  select(Women,Men)

sports_by_gender <- all_sports %>%
  pivot_wider(names_from = gender, values_from = participants) %>%
  arrange(Sport)

total_sports %>%
  kable(escape = F, booktabs = T, format = "latex", align = "c", caption = "Number of total sports offered, by gender") %>%
  kable_styling(latex_options = c("HOLD_position","repeat_header"), full_width = TRUE) 

sports_by_gender %>%
  adorn_totals() %>%
  mutate(across(where(is.numeric), ~as.character(.x))) %>%
  mutate(across(everything(), ~ifelse(is.na(.x), "\u2013",.x))) %>%
  kable(escape = F, booktabs = T, format = "latex", align = "c", caption = "Number of students participating in sports, by gender") %>%
  kable_styling(latex_options = c("HOLD_position","repeat_header"), full_width = TRUE) %>%
  row_spec(nrow(sports_by_gender), hline_after = T, extra_latex_after = "%") %>%
  row_spec(nrow(sports_by_gender)+1,bold=T) 

```

## II. Salary Snapshot

```{r}
#| echo: false
#| output: true

plot_dta <- dta %>%
  select(starts_with(c("STUDENTAID","RECRUITEXP","ASCOACH_SAL_FTE","HDCOACH_SAL_FTE")) & !ends_with(c("COED","TOTAL"))) %>%
  pivot_longer(
    cols = everything(),
    names_to = c("type","gender"),
    values_to = "expense",
    names_pattern = "(STUDENTAID|RECRUITEXP|HDCOACH_SALARY|ASCOACH_SAL_FTE|HDCOACH_SAL_FTE)_(.*)"
  ) 

coach_sal <- plot_dta %>%
  filter(grepl("SAL_FTE",type, fixed = TRUE)) %>%
  mutate(type = case_when(grepl("ASCOACH",type,fixed = TRUE) ~ "Assistant coach",
                          TRUE                               ~ "Head coach"),
         gender = case_when(gender == "WOMN" ~ "Women's teams",
                            TRUE             ~ "Men's teams")) %>%
  mutate(type = factor(type, levels = c("Head coach","Assistant coach")))
  
ggplot(data = coach_sal, aes(x = type, y = expense, fill = gender)) +
  geom_bar(stat = "identity", position = "dodge", color = "black") +
  geom_text(aes(label = scales::dollar(expense)), position = position_dodge(width = 0.9), vjust = -1) +
  theme_minimal() +
  labs(x = "",
       y = "Average full-time salary") +
  ggtitle("Figure 1: Average full-time coaching salary, \nby team gender and coach type") +
  theme(plot.title = element_text(hjust = 0.5),
        legend.title = element_blank(),
        legend.key = element_blank(),
        legend.background = element_rect(color = NA),
        legend.position="bottom") +
  scale_fill_brewer(palette = "Accent") +
  scale_y_continuous(labels = scales::dollar_format(prefix="$"), expand = c(0.2,0))
  
```
## References

`r glue("U.S. Department of Education, Office of Postsecondary Education, Equity in Athletics Disclosure Act (EADA) survey, {sy}.")`